---
title: "CMTH642_Assignment_02"
author: "Tusaif Azmat"
date: "27/03/2022"
output:
  word_document: default
  html_document: default
  pdf_document: default
---
---
## Assignment 2 (10%) 
## Tusaif Azmat 
## Student#500660278 
---

## R Markdown


QUESTIONS
1. Check the datatypes of the attributes. (3 points)

```{r}
wine_df<-read.csv("C:/Users/Zanara/Documents/Ryerson/Winter2022/CMTH642/CMTH642_winter2022/A2/A2/winequality-white.csv",header= T,sep = ";")

head(wine_df)
#You could see the data types of each attribute under column names all double except one integer, all numeric values 
```
```{r}
sapply(wine_df, class)
#You could see the data types of each attribute
```


2. Are there any missing values in the dataset? (4 points)
```{r}
which(is.na(wine_df))

#There are no missing values in the dataset as seen below.
```


3. What is the correlation between the attributes other than Quality? (10 points)
```{r}
cor(wine_df[-12])
#A correlation is a number between -1 and +1 that measures the degree of association between two Attributes (call them X and Y). A positive value for the correlation implies a positive association. In this case large values of X tend to be associated with large values of Y and small values of X tend to be associated with small values of Y. A negative value for the correlation implies a negative or inverse association. In this case large values of X tend to be associated with small values of Y and vice versa.
#Following are the correlation values...of all attributes except quality.
```


4. Graph the frequency distribution of wine quality by using Quality. (10 points)
```{r}
hist(wine_df$quality, main="Wine Quality Distribution")
```


5. Reduce the levels of rating for quality to two levels as Pass and Fail. Assign the levels of 3, 4 and 5 to level Fail; and 6, 7, 8 and 9 to level Pass. (10 points)

```{r}
wine_df$quality<-as.factor(ifelse(wine_df$quality > 5,1,0))


table(wine_df$quality)

#fail = 0
#pass = 1
#I use zero for fail and one for pass and you could see below the values
```


6. Normalize the data set. (12 points)
```{r}
normalize <- function(x) {
  return ((x-min(x))/(max(x)-min(x)))
}
#new normalized dataset created below 
wine_df_new<-wine_df
wine_df_new[,-12] <- sapply(wine_df_new[,-12], normalize)
summary(wine_df_new)

#following are the normalized dataset values
```


7. Divide the dataset to training and test sets. (12 points)
```{r}
#I use the 70 30 split of dataset training and test sets
train_index = sample(1:nrow(wine_df_new),0.7*nrow(wine_df_new))
train.set= wine_df_new[train_index,]
test.set= wine_df_new[-train_index,]

```


8. Use the Logistic Regression algorithm to predict the quality of wine using its attributes. (12 points)

```{r}
LR_model<-glm(formula =quality~.,data=train.set,family = "binomial")
summary(LR_model)
# Number of Fisher Scoring iterations: This is just a measure of how long it took to fit your model. You can safely ignore it.
```

```{r}

pred=predict(LR_model,type ='response',newdata = test.set)

predicted.quality<-ifelse(pred>=0.717189,1,0)

hist(pred)
```

```{r}
# I decided to use the median to predict if the model can tell whether the quality of a given wine will pass. 
# Due to an imbalanced dataset we have clear problems with skewness in the predicted variable of our model. 
# This would affect the performance of the model by making it less accurate at its prediction. 
summary(pred)
```

9. Display the confusion matrix to evaluate the model performance. (12 points)
```{r}
c.matrix<-table(actual=test.set$quality,pred=predicted.quality)
c.matrix
#The results are not quite convinced
```


10. Evaluate the model performance by computing Accuracy, Sensitivity and Specificity. (15 points)
```{r}
TP=c.matrix["0","0"]
FP=c.matrix["1","0"]
FN=c.matrix["0","1"]
TN=c.matrix["1","1"]

#Accuracy
Accuracy=(TP+TN)/(TP+FN+FP+TN)
writeLines("Accuracy")
Accuracy

#Sensitivity

Sensitivity=TP/(TP+FN)
writeLines("Sensitivity")
Sensitivity

#Specificity

Specificity=TN/(TN+FP)
writeLines("Specificity")
Specificity
```

I obtained accuracy of the model 68.64 percent, Sensitivity of 82.07% and Specificity of 62.25% .

#This is the end of Assignment 2
## R Markdown File