---
title: 'CMTH 642 Data Analytics: Advanced Methods'
output:
  pdf_document: default
  html_document: default
  word_document: default
---
<center> <h1> Assignment 1 (10%) </h1> </center>
<center>  <h3> Tusaif Azmat </h2> </center>
<center> <h3> CMTH642-DJ0 500660278 </h2> </center>
---


#### 1. Read the csv files in the folder. (4 points)
```{r}
df_usda_macro <- read.csv("USDA_Macronutrients.csv")
df_usda_micro <- read.csv("USDA_Micronutrients.csv")
```

#### 2. Merge the data frames using the variable "ID". Name the Merged Data Frame "USDA". (4 points)
```{r}
USDA <- merge(df_usda_macro, df_usda_micro, by='ID')
summary(USDA)
```

#### 3. Check the datatypes of the attributes. Delete the commas in the Sodium and Potasium records. Assign Sodium and Potasium as numeric data types. (4 points)
```{r}
sapply(USDA, class)

USDA$Sodium <- gsub(',','',USDA$Sodium)
USDA$Potassium <- gsub(',','', USDA$Potassium)

USDA$Sodium = as.numeric(USDA$Sodium)
USDA$Potassium = as.numeric(USDA$Potassium)

```

#### 4. Remove records (rows) with missing values in more than 4 attributes (columns). How many records remain in the data frame? (4 points)
```{r}
USDA.nacount <- apply(USDA,1, function(x) sum(is.na(x)))

USDATrim <- USDA[USDA.nacount <= 4,]
nrow(USDATrim)
```

#### 5. For records with missing values for Sugar, Vitamin E and Vitamin D, replace missing values with mean value for the respective variable. (4 points)
```{r}
USDA$Sugar[is.na(USDA$Sugar)] = mean(USDA$Sugar[!is.na(USDA$Sugar)])

USDA$VitaminE[is.na(USDA$VitaminE)] = mean(USDA$VitaminE[!is.na(USDA$VitaminE)])

USDA$VitaminD[is.na(USDA$VitaminD)] = mean(USDA$VitaminD[!is.na(USDA$VitaminD)])
```

#### 6. With a single line of code, remove all remaining records with missing values. Name the new Data Frame "USDAclean". How many records remain in the data frame? (5 points)
```{r}
USDAclean = USDA[complete.cases(USDA),]
# 6310 records remain
cat(nrow(USDAclean), " records remain in a the data frame.")
```

#### 7. Which food has the highest sodium level? (5 points)
```{r}
USDAclean$Description[which.max(USDAclean$Sodium)]
# Table Salt with ID 2047 has the highest Sodium of 38758
```

#### 8. Create a histogram of Vitamin C distribution in foods. (5 points)
```{r}

hist(USDAclean$VitaminC, xlim = c(1, 100), breaks = 100, xlab = "Vitamin C", ylab= "Frequency of each level", main="Vitamin C Distribution in Foods")

```

#### 9. Create one boxplot to illustrate the distribution of values for TotalFat, Protein and Carbohydrate. (5 points)
```{r}

TPC <- list(USDAclean$TotalFat, USDAclean$Protein, USDAclean$Carbohydrate)
names(TPC) <- c("TotalFat", "Protein", "Carbohydrate")
boxplot(TPC, main="Distribution of Values for TotalFat, Protein and Carbohydrate")

```

#### 10. Create a scatterplot to illustrate the relationship between a food's TotalFat content and its Calorie content. (5 points)
```{r}

plot(USDAclean$Calories~USDAclean$TotalFat, main="Relationship between Food's TotalFat and Calorie content", ylab="Calories", xlab="Total Fat")

```

#### 11. Add a variable to the data frame that takes value 1 if the food has higher sodium than average, 0 otherwise. Call this variable HighSodium. Do the same for High Calories, High Protein, High Sugar, and High Fat. How many foods have both high sodium and high fat? (5 points)
```{r}
USDAclean$HighSodium = 0
USDAclean$HighSodium[USDAclean$Sodium > mean(USDAclean$Sodium)] = 1

USDAclean$HighCalories = 0
USDAclean$HighCalories[USDAclean$Calories > mean(USDAclean$Calories)] = 1

USDAclean$HighProtein = 0
USDAclean$HighProtein[USDAclean$Protein > mean(USDAclean$Protein)] = 1

USDAclean$HighSugar = 0
USDAclean$HighSugar[USDAclean$Sugar > mean(USDAclean$Sugar)] = 1

USDAclean$HighFat = 0
USDAclean$HighFat[USDAclean$TotalFat > mean(USDAclean$TotalFat)] = 1

cat(sum(apply(USDAclean[c("HighSodium", "HighFat")], 1, function(x) sum(x) == 2)), "foods have both high sodium and high fat.")

# 644 foods high sodium and high fat.
```

#### 12. Calculate the average amount of iron, for high and low protein foods. (5 points)
```{r}
MeanProteinIron <- aggregate(USDAclean$Iron,list(USDAclean$HighProtein),FUN = mean) 
colnames(MeanProteinIron) <- c("low(0)/high(1) protein","AVG amount")
head(MeanProteinIron)

# 3.069541 Iron for High Protein
# 2.696634 Iron for Low Protein
```

#### 13. Create a function for a "HealthCheck" program to detect unhealthy foods. Use the algorithm flowchart below as a basis. (5 points)
```{r}
require(jpeg)
img<-readJPEG("HealthCheck.jpg")
plot(1:4, ty = 'n', ann = F, xaxt = 'n', yaxt = 'n')
rasterImage(img,1,1,4,4)
```
```{r}
HealthCheck <- function (sodium,sugar,fat) {
  ifelse (sodium==0, "Pass", ifelse (sugar==0, "Pass", ifelse (fat==0, "Pass", "Fail")))
}

```

#### 14. Add a new variable called HealthCheck to the data frame using the output of the function. (5 points)
```{r}
USDAclean$HealthCheck = HealthCheck(USDAclean$HighSodium, USDAclean$HighSugar, USDAclean$HighFat)
```

#### 15. How many foods in the USDAclean data frame fail the HealthCheck? (5 points)
```{r}
sum(USDAclean$HealthCheck == "Fail",na.rm = TRUE)
# 237 food fail Health check
```

#### 16. Visualize the correlation among Calories, Protein, Total Fat, Carbohydrate, Sodium and Cholesterol. (5 points)
```{r}
cor(USDAclean[3:8])
```

#### 17. Is the correlation between Calories and Total Fat statistically significant? Why? (5 points)
```{r}
cor.test(USDAclean$Calories,USDAclean$TotalFat)

#Yes, The correlation between Calories and Total Fat is statistically significant.As described below, the p-value, 2.2e-16, is less than the confidence level of 0.05.The relationship would be described as a high positive relationship.
```

#### 18. Create a Linear Regression Model, using Calories as the dependent variable Protein, Total Fat, Carbohydrate, Sodium and Cholesterol as the independent variables. (4 points)
```{r}
lm_USDA <- lm(USDAclean$Calories ~ USDAclean$Protein + USDAclean$TotalFat+
               USDAclean$Carbohydrate + USDAclean$Sodium + USDAclean$Cholesterol)
summary(lm_USDA)

# y = 3.9882753 + 3.9891994(Protein) + 8.7716980(TotalFat) + 3.7432001(Carbohydrate) + 0.0003383(Sodium) + 0.0110138(Cholesterol)

```

#### 19. Which independent variable is the least significant? Why? (4 points)
```{r}
lm_Anova_USDA <- anova(lm_USDA)

lm_Anova_USDA

# Sodium is the least significant variable. We can interpret this from the analysis of Variance Table. The p-value for sodium is 0.1378, which is not very significant, especially compared to the p-values of the other variables, which are all much smaller, less than 2e-16.
```

#### 20. Create a new model by using only the significant independent variables. (4 points)
```{r}
lm_USDA_new <- lm(USDAclean$Calories ~ USDAclean$Protein + USDAclean$TotalFat+
               USDAclean$Carbohydrate + USDAclean$Cholesterol)
summary(lm_USDA_new)

lm_Anova_USDA_new <- anova(lm_USDA_new)

lm_Anova_USDA_new
```

#### 21. A new product is just produced with the following data: Protein=0.1, TotalFat=37, Carbohydrate=400, Cholesterol=75, Sugar=NA, Calcium=35, Iron=NA, Potassium=35, VitaminC=10,	VitaminE=NA, VitaminD=NA. Based on the new model you created, what is the predicted value for Calories? (4 points)
```{r}
New_Product <- data.frame(Protein=0.1, TotalFat=37, Carbohydrate=400, Sodium=440, Cholesterol=75, Sugar=NA, Calcium=35, Iron=NA, Potassium=35, VitaminC=10,	VitaminE=NA, VitaminD=NA)


Predicted_Calories_value <- 3.9882753 + 3.9891994*New_Product$Protein +
8.7716980*New_Product$TotalFat + 3.7432001*New_Product$Carbohydrate +
0.0003383*New_Product$Sodium + 0.0110138*New_Product$Cholesterol

Predicted_Calories_value

#The predicted value would be 1827.195
```

#### 22. If the Carbohydrate amount increases from 400 to 40000 (10000% increase), how much change will occur on Calories in percent? Explain why? (4 points)
```{r}
Predicted_Calories_Increased <- 3.9882753 + 3.9891994*New_Product$Protein +
8.7716980*New_Product$TotalFat + 3.7432001*New_Product$Carbohydrate +
0.0003383*44440 + 0.0110138*New_Product$Cholesterol

Predicted_Calories_Increased

Change_in_Calories <- (44440-440)*0.0003383
Percentage_of_Change <- (Change_in_Calories/Predicted_Calories_value)* 100
Percentage_of_Change

# If the value of Sodium increased from 440 to 44440, the value of Calories would change by 14.8852. 
# This represents a 0.81% change in the value of Calories from when Sodium was equal to 440.
# To get this result we multiply the difference in the Sodium value from before to after (44440-440 = 44000) by the coefficient for Sodium from the model, which is 0.0003383. 
# We use this value in our regression calculation. The coefficient describes the change in the dependent variable for each unit of change in the Sodium variable.
```


This is the end of Assignment 1

Ceni Babaoglu, PhD
